---
title: "Graphing Using whitewater"
author: "Josh Erickson"
date: '2023-01-28'
slug: []
categories:
- R
- Hydrology
tags:
- dataRetrieval
- Hydrology
- R
- https://github.com/USGS-R/dataRetrieval
subtitle: ''
excerpt: ''
draft: yes
series: null
layout: single
---


```{r, message=FALSE, warning=FALSE, error = F, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")),fig.width = 10, fig.height = 5,
                      warning = FALSE, message = FALSE, error = FALSE)
```

## Intro  
whitewater was originally developed for parallel processing but has always had some data wrangling and tranformations to help visualise USGS NWIS data quickly. In this post we'll use the advantages of whitewater to look at a site graphically a few different ways (tiled, ). Then we can take these functional programming advantages and use within a reactive framework like R shiny.  


## Get Data  

Let's load the packages. We'll use {ggfx} to give it an extra effects but you don't necessarily need it.  


```{r}
library(whitewater)
library(dplyr)
library(ggplot2)
library(ggfx)
```

Now let's get a sites daily value. We'll get a couple different flow regimes (snowmelt vs precipitation).  

```{r}
yaak_dv <- ww_dvUSGS('12304500',
                     wy_month = 10)

elwah_dv <- ww_dvUSGS('12045500',
                                wy_month = 10)
```

## Raster Graph  

This is an approach I saw during a presentation and can be accessed as well through the [Raster-Hydrograph](https://waterwatch.usgs.gov/index.php?id=wwchart_rastergraph) builder online.  

Below is the code to create raster-hydrograph plot using {ggplot2}. What's nice is the `wy_doy` and `month_abb` that whitewater returns. This let's us stack the water years and the day of the year like a raster or tile. This is essentially a heatmap by water and day of year filled by flow value.    

```{r}
raster_hydrograph <- function(daily_values) {
  
xbreaks <- c(1, 32,  62,  93, 124, 152, 183, 213, 244, 274, 305, 336)

dup_labels <- function(x) daily_values$month_abb[match(x, daily_values$wy_doy)]

daily_values %>% 
  ggplot(aes(wy_doy, wy)) + 
  with_outer_glow(geom_tile(aes(fill = Flow))) +
  scale_y_continuous( breaks = seq(min(daily_values$wy),
                                  max(daily_values$wy),
                                  by = 10),
                     name = 'Water Year') +
  scale_x_continuous(breaks = xbreaks,
                     name = 'Day of Year',
                     sec.axis = dup_axis(labels = dup_labels, name = NULL),
                     expand = c(0,0)) + 
  scale_fill_gradientn(colors = hcl.colors(11, palette = 'Spectral'),
                       labels = scales::comma,
                       name = 'Discharge (cfs)',
                       trans = 'log') +
  labs(title = 'Raster-Hydrograph of Daily Discharge (cfs)',
       subtitle = paste0('at USGS site ', daily_values[1,]$site_no,
                         ' ',
                         daily_values[1,]$Station)) + 
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 8),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_rect(fill = NA),
        panel.ontop = T, 
        axis.ticks.x = element_blank(),
        axis.text.x.bottom = element_text(vjust = 5.5,face = 'bold'), 
        axis.text.x.top = element_text(hjust = -.45, vjust = -5.5, face = 'bold'))
}
```

Now we can plot any station from the `ww_dvUSGS()` functions. Notice how different the two graphs are! This is a reflection of the different flow regimes and how distribution and timing of runoff matters.  


```{r}
raster_hydrograph(yaak_dv)
raster_hydrograph(elwah_dv)
```


## Hysteresis  

Another way to view these flow regime differences is through a lagged version of itself. In other words, a bivariate plot of flow and its lagged flow. In a basic sense the flow will make clockwise circles. The width of these circles can help understand the response of the system, i.e. wider meaning flashier and narrow meaning slow steady increases.  

To do this we'll need to create som lagged versions of the daily value tables.  

```{r}
lagged_yaak <- yaak_dv %>% mutate(flow_lag = lag(Flow, 1))
lagged_elwah <- elwah_dv %>% mutate(flow_lag = lag(Flow, 1))

```

```{r, echo=F}
library(patchwork)
p1 <- lagged_elwah %>% 
  filter(wy == 2022) %>% 
  ggplot(aes(Flow, flow_lag)) + 
  geom_path(aes(color = wy_doy), 
            arrow = ggplot2::arrow(
        length = ggplot2::unit(1.5, "mm"),
        ends = "last"
      ), show.legend = F) +
  labs(y = 'Lagged Discharge (1 day)') +
  geom_abline(slope = 1, intercept = 0) + 
  facet_wrap(~Station, scales = 'free') +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_bw() +
  theme(strip.text = element_text(size = 6))

p2 <- lagged_yaak %>% 
  filter(wy == 2022) %>% 
  ggplot(aes(Flow, flow_lag)) + 
  geom_path(aes(color = wy_doy), 
            arrow = ggplot2::arrow(
        length = ggplot2::unit(1.5, "mm"),
        ends = "last"
      ), show.legend = F) +
  labs(y = 'Lagged Discharge (1 day)') +
  geom_abline(slope = 1, intercept = 0) + 
  facet_wrap(~Station, scales = 'free') +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_bw()+
  theme(strip.text = element_text(size = 6))

(p1|p2) + plot_annotation(title = 'Comparing Water Year 2022 Discharge with Lagged Discharge')
```


Now let's create a function like we did before to look at all the months.  

```{r}

hysteresis_plot <- function(daily_values, lag_days = 1){
  
lagged_df <- daily_values %>% mutate(flow_lag = lag(Flow, lag_days))

lagged_df <- lagged_df %>% filter(!is.na(flow_lag))

lagged_df_final <- lagged_df %>% group_by(month_abb) %>% 
             tidyr::nest() %>% 
             mutate(model = purrr::map(data, ~lm(.$flow_lag~.$Flow, data = .)),
                    pred = purrr::map2(data, model, ~as.numeric(predict(.y, .x)))) %>% 
             select(data, pred) %>% 
             tidyr::unnest(c('data','pred'))

  lagged_df_final %>% 
  ggplot(aes(Flow, flow_lag)) + 
  geom_path(aes(color = wy), 
            alpha = 0.1,
            arrow = ggplot2::arrow(
              length = ggplot2::unit(1.5, "mm"),
              ends = "last"
            )) +
  geom_abline(slope = 1, intercept = 0) + 
  stat_smooth(geom = 'line',
              method = 'lm',
              linetype = 2, aes(Flow, pred)) +
  scale_x_continuous(labels = scales::comma) + 
  scale_y_continuous(labels = scales::comma) + 
  labs(y = 'Lagged Flow') + 
  facet_wrap(~month_abb, scales = 'free') +
  theme_bw() +
  theme(axis.text = element_text(size = 5))

}

```

```{r, message=F, warning=F}
hysteresis_plot(yaak_dv)
hysteresis_plot(elwah_dv)
```


```{r, message=F, warning=F}

tob_dv <- ww_dvUSGS('12301300')

raster_hydrograph(tob_dv)
hysteresis_plot(tob_dv)
```

